---
title: "Practical session on ISwR::cystfibr dataset"
author: "Magali Richard & Florent Chuffart & Nicolas Glade"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---



```{r, echo=FALSE, eval=TRUE, label="loading libraries"}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide")
```

# Overview

The goal of this practical session is to explore data analysis methods based on variance and correlation. 

Use will analyse `ISwR::cystfibr`  dataset and use k-means, PCA and ANOVA methods.  

```
# ?ISwR::cystfibr

cystfibr                 package:ISwR                  R Documentation

Cystic fibrosis lung function data

Description:

     The ‘cystfibr’ data frame has 25 rows and 10 columns.  It contains
     lung function data for cystic fibrosis patients (7-23 years old).

Usage:

     cystfibr
     
Format:

     This data frame contains the following columns:

     ‘age’ a numeric vector, age in years.

     ‘sex’ a numeric vector code, 0: male, 1:female.

     ‘height’ a numeric vector, height (cm).

     ‘weight’ a numeric vector, weight (kg).

     ‘bmp’ a numeric vector, body mass (% of normal).

     ‘fev1’ a numeric vector, forced expiratory volume.

     ‘rv’ a numeric vector, residual volume.

     ‘frc’ a numeric vector, functional residual capacity.

     ‘tlc’ a numeric vector, total lung capacity.

     ‘pemax’ a numeric vector, maximum expiratory pressure.

Source:

     D.G. Altman (1991), _Practical Statistics for Medical Research_,
     Table 12.11, Chapman & Hall.

References:

     O’Neill et al. (1983), The effects of chronic hyperinflation,
     nutritional status, and posture on respiratory muscle strength in
     cystic fibrosis, _Am. Rev. Respir. Dis.,_ 128:1051-1054.

```

We do not consider `sex` variable.

We considere `pemax` as disease severity indicator.

We try to explain `pemax` according to other variables.


```{r echo=TRUE, results="verbatim"}
d = ISwR::cystfibr
head(d)
dim(d)
# removing sex
d = d[,-2]
```


# Methods

## Correlation

```{r echo=TRUE, results="verbatim", fig.width=9, fig.height=9}
pairs(d, gap=0)
lattice::levelplot((cor(d)))
```

## K-means

```{r echo=TRUE, results="verbatim", fig.width=9, fig.height=9}
centers = 2
k = kmeans(d, centers=centers);  
# pairs(d, gap=0, col=k$cluster, main=paste0("centers=", centers))
k

cs = rep(2:20, each=100)
res = sapply(cs, function(c) {
  k = kmeans(d, centers=c);  
  ret = k$betweenss / k$totss
  return(ret)
})
plot(jitter(cs), jitter(res))
```


## PCA

```{r}
# mat = x[,-9]
mat = d
pca = prcomp(mat, scale=TRUE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100

# layout(matrix(1:6,2), respect=TRUE)
layout(matrix(1:2,1), respect=TRUE)
barplot(p)

for (i in 1:5) {
  j = i+1
  plot(pca$x[,i], pca$x[,j], xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"), ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"), pch=16)  
  scale_factor = min(abs(c(min(c(pca$x[,i], pca$x[,j])), max(c(pca$x[,i], pca$x[,j])))))  
  # scale_factor = min(abs(c(max(min(pca$x[,i]), min(pca$x[,j])), min(max(pca$x[,i]), max(pca$x[,j])))))
  plotrix::draw.ellipse(0,0,scale_factor,scale_factor, lty=2, border="grey")
  arrows(0,0,pca$rotation[,i]*scale_factor, pca$rotation[,j]*scale_factor, col="grey")
  text(pca$rotation[,i]*scale_factor, pca$rotation[,j]*scale_factor, rownames(pca$rotation))
}
```

## Linear Regression 


### Variable selection

step backward

$$AIC=-log(L) + 2k$$


```{r echp=TRUE, results="verbatim"}
m = lm(pemax~age+height+weight+bmp+fev1+rv+frc+tlc, d)
s = step(m)
shapiro.test(s$res)
summary(s)


m = lm(formula = pemax ~ weight + bmp + fev1 + rv, data = d)
summary(m)$r.squared
extractAIC(m)

m1 = lm(formula = pemax ~ age + bmp + fev1 + rv, data = d)
summary(m1)$r.squared
extractAIC(m1)

m3 = lm(formula = pemax ~ weight + bmp + fev1 + frc, data = d)
summary(m3)$r.squared
extractAIC(m3)

m4 = lm(formula = pemax ~ weight + bmp + fev1, data = d)
summary(m4)$r.squared
extractAIC(m4)
anova(m4)

d$bmp_bin = d$bmp < 80
m = lm(formula = pemax ~ weight + bmp_bin + fev1 + rv, data = d)
summary(m)$r.squared
extractAIC(m)








m = lm(formula = pemax ~ weight + fev1 + bmp + rv, data = d)
anova(m)

m = lm(formula = pemax ~ weight + rv + fev1 + bmp, data = d)
anova(m)

```



# Discussion

...


# Session Information

```{r, results="verbatim"}
sessionInfo()
```

